Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-01 text/back_end_KS.qmd
methods

Our study utilizes two main data sources based on each research question. The first research question is focused on understanding the socioeconomic and racial variation between Geomarkets in metropolitan areas and how such variation changes over time. To answer this question, we use census tract data from the U.S. Census Bureau. Utilizing a small geographic area is important for implementing a spatial merge between Census data and Geomarket shape files (described below). Census tract is the smallest geographic area for which measures of both race/ethnicity and socioeconomic characteristics are available. Census tracts are statistical subdivisions of a county designed to capture relatively homogeneous demographic communities that are generally considered to be more accurate for population analyses than zip codes, which are spatial boundaries drawn to demarcate route efficiency for postal services \[[CITE](https://www.census.gov/newsroom/blogs/random-samplings/2014/07/understanding-geographic-relationships-counties-places-tracts-and-more.html)\] .

@@

@@

**Research question 1, variables**. RQ1 asks, what is the socioeconomic and racial variation between Geomarkets in metropolitan areas? And, how does this variation change over time? We created measures of race and ethnicity, specifically the number and proportion of people (all ages) in the following categories: White, non-Hispanic; Black, non-Hispanic; Asian, non-Hispanic; Pacific Islander, non-Hispanic; American Indian and Alaskan Native, non-Hispanic; and two+ races, non-Hispanic. Measures of XXX were not available for 1980 Decennial Census. We also created socioeconomic measures that approxmiate the predictors of market segment identified by @RN4982. These variables are median household income, mean household income, percent of households below the poverty line, and people age 25+ with a BA.[^4] All measures were created at the census tract-level and the Geomarket-level by aggregating tract-level data to Geomarkets.

[^4]: STATE THAT MEAN HOUSEHOLD INCOME CREATED BY DIVIDING AGGREGATE INCOME BY NUMBER OF PEOPLE. NEED FIX THIS MEASURE THO.

**Research question 2, data**. Geomarket is a search filter available in the College Board Student Search Service product. RQ2 asks, how does the racial and socioeconomic composition of included versus included prospects vary when student list purchases filter on particular geomarkets? We answer this question using actual student lists purchased by public universities.

In February 2020, we began issuing public records requests to public universities about student lists purchased from 2016 through 2020. We narrowed the scope of our request to student lists purchased from College Board, ACT, and the National Research Center for College and University Admissions (NRCCUA), the three largest student list vendors at the time. For each student list purchased from 2016 through 2020, we requested two related pieces of data: (1) the order summary, which specifies search criteria for the student list purchase; and (2) the de-identified prospect-level list produced from the search criteria.

The data collection sample for this project was all public universities in CA, IL, MN, and TX. Utilizing public records requests to obtain public records is a painstaking process. Initially, the majority of universities did not respond to our request or denied our request. Subsequently, we obtained pro bono representation from four law firms, which substantially increased the success of data collection. We collected data from Arizona State University and Northern Arizona University, because we were able to obtain pro-bono legal representation for these two universities. However, we were unable to obtain legal representation for TX.

Even with firm representation, data collection remained difficult. Some universities provided records that were not usable for quantitative analyses (e.g., summary statistics across multiple orders; or data did not contain important fields). Some universities did not provide records based on legitimate grounds (e.g., data not in university possession; not required to create records that do not currently exist). We learned that many universities outsourced student list purchases to a third-party consulting firm. Unfortunately, we were rarely able to obtain usable data from these universities. A small number of universities denied requests based on potentially questionable legal rationale, but we lacked the resources to litigate.

This article analyzes student lists purchased from College Board. We exclude student list purchases by MN public universities from this report because Minnesota is predominantly an "ACT state." 14 public universities provided PDF order summaries containing the search criteria for 830 student list purchases. We utilized the XXXX Python/R package to convert these order summaries into tabular data. About 16% of student list purchases utilized the Geomarket filter.

We received both (a) the order summary data and (b) the de-identified prospect-level student list data for 414 orders associated with 2,549,085 prospects. We draw from these data to answer RQ2. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase.[^5]. Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

[^5]: ADD CAVEAT THAT PURCHASES EXCLUDE PROSPECT PROFILES THAT WERE PREVIOUSLY PURCHASED

In particular, we answer RQ2 by analyzing particular student list purchases that did *not* utilize the Geomarket search filter, but did purchase prospects from all Geomarkets in a particular metropolitan area. For each prospect we know their home zip code and, therefore, which Geomarket they belong to. We answer RQ2 by simulating which prospective students would have been included or excluded from the student list purchase had the purchase filtered on partiulcar Geomarkets.

**Research question 2, variables**. The variables of interest for RQ2 the characteristics of prospective students who would have been included/excluded by our simulations and the characteristics of their schools and neighborhoods. Prospect characteristics are derived from the pretest questionnaire administered to College Board test-takers. The de-identified student lists we received contain a standard subset of variables (e.g., HS code, HS GPA range, intended major). We create detailed measures of race and ethnicity. Information about student socioeconomic status is limited. However, we create a measure of whether the prospect is first-generation college student. We can also measure the socioeconomic characteristics of the students' home zip code and characteristics of their high school \[EXPAND OR DELETE THIS SENTENCE DEPENDING ON WHETHER WE USE THESE DATA\].

@@

**Research design**. We utilize a multiple, quantitative case study design in which metropolitan areas are cases.

Following XXX \[CITE\], we focus on a small number \[HOW MANY\] of metropolitan areas in order to provide sufficient analytic depth while also situating each case within the context of historic segregation and contemporary gentrification. Online Appendix XXX presents results for a larger number of metropolitan areas.

Choice of metropolitan areas is informed by several factors. Based on our read of @RN4982, we are interested in "major metropolitan area\[s\] \ldots composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs" (p. 11-12). Building on our conceptual framework, we are interested in racial and socioeconomic inequality between geomarkets within a metropolitan area (RQ1) and how this inequality contributes to inequality in which prospective students are targeted by student list purchases (RQ2). \[DISCUSSION OF WHICH CASES TO SELECT. KARINA\]. Choice of metropolitan areas is informed by data availability. Although Census data are available for all metropolitan areas (RQ1) we do not have good candidate student list purchases for all metropolitan areas (RQ2). \[SAY SOMETHING ABOUT WHAT IS A GOOD CANDIDATE STUDENT LIST PURCHASE -- PURCHASE WITH NOT TOO MANY CRITERIA THAT INCLUDE ALL GEOMARKETS IN A METRO\].

>>>>

**RQ1**. Our study utilizes two main data sources based on each research question. The first research question is focused on understanding the socioeconomic and racial variation between Geomarkets in metropolitan areas and how such variation changes over time. To answer this question, we use census tract data from the U.S. Census Bureau. Utilizing a small geographic area is important for implementing a spatial merge between Census data and Geomarket shape files (described below). Census tract is the smallest geographic area for which measures of both race/ethnicity and socioeconomic characteristics are available. Census tracts are statistical subdivisions of a county designed to capture relatively homogeneous demographic communities that are generally considered to be more accurate for population analyses than zip codes, which are spatial boundaries drawn to demarcate route efficiency for postal services \[[CITE](https://www.census.gov/newsroom/blogs/random-samplings/2014/07/understanding-geographic-relationships-counties-places-tracts-and-more.html)\] .

@@

Across these data sources, we created measures of race and ethnicity, specifically the number and proportion of the census-tract population in the following categories: White, non-Hispanic; Black, non-Hispanic; Asian, non-Hispanic; Pacific Islander, non-Hispanic; American Indian and Alaskan Native, non-Hispanic; and two or more races, non-Hispanic. Measures of Asian, American Indian and Alaskan Native, two or more races, and Native Hawaiian and Pacific Islander (all non-Hispanic) were not available for the 1980 Decennial Census. We also created socioeconomic measures that approximate the predictors of market segment identified by @RN4982. These variables are median household income, percent of households below the poverty line, and people age 25+ with a BA. All measures were created at the census tract-level and the Geomarket-level by aggregating tract-level data to Geomarkets.

@@

**RQ2**. Our study's second research question asks how does the racial and socioeconomic composition of included versus included prospects vary when student list purchases filter on Geomarkets. We answer this question using actual student lists purchased by public universities. In February 2020, we began issuing public records requests to public universities about student lists purchased from 2016 through 2020. We narrowed the scope of our request to student lists purchased from College Board, ACT, and the National Research Center for College and University Admissions (NRCCUA), the three largest student list vendors at the time. For each student list purchased from 2016 through 2020, we requested two related pieces of data: (1) the order summary, which specifies search criteria for the student list purchase; and (2) the de-identified prospect-level list produced from the search criteria.

The data collection sample for this project was all public universities in California, Illinois, Minnesota, and Texas. Utilizing public records requests to obtain public records is a painstaking process. Initially, the majority of universities did not respond or denied our request. Subsequently, we obtained pro bono representation from four law firms, which substantially increased the success of data collection. We also collected data from Arizona State University and Northern Arizona University, because we were able to obtain pro-bono legal representation for these two universities. However, we were unable to obtain legal representation for Texas.

Even with firm representation, data collection remained difficult. Some universities provided records that were not usable for quantitative analyses (e.g., summary statistics across multiple orders; or data did not contain important fields). Some universities did not provide records based on legitimate legal grounds (e.g., data not in university possession; not required to create records that do not currently exist). We learned that many universities outsourced student list purchases to third-party consulting firms. Unfortunately, we were rarely able to obtain usable data from these universities. A small number of universities denied requests based on potentially questionable legal rationale, but we lacked the resources to litigate.

The primary data source for our second research question is therefore student lists purchased from College Board for which we received both (a) the order summary data and (b) the de-identified prospect-level student list data. This sample includes a total of 414 orders associated with 2,549,085 prospects made by 14 \[KS CHECK THIS\] public universities located in California, Illinois, Minnesota, and Arizona \[KS Does this still apply?\]. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase.[^4] The resulting student lists could hypothetically extend to any university placing student lists orders based on the same filter criteria. In other words, any observed patterns in the findings below would remain constant for any university selecting the same filters. Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

[^4]: ADD CAVEAT THAT PURCHASES EXCLUDE PROSPECT PROFILES THAT WERE PREVIOUSLY PURCHASED

In particular, we answer the second research question by analyzing particular student list purchases that did *not* utilize the Geomarket search filter, but did purchase prospects from all Geomarkets in a particular metropolitan area. For each prospect we know their home zip code and, therefore, which Geomarket they belong to. We therefore simulate which prospective students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets.

The variables of interest for student list data are the characteristics of prospects who would have been included versus excluded by our simulations by Geomarket filter. Prospect characteristics are derived from the pretest questionnaire administered to College Board test-takers that are included in the de-identified student lists we received (e.g., HS code, HS GPA range, intended major). College Board asks students separate questions about ethnicity (Cuban, Mexican, Puerto Rican, other Hispanic, non-Hispanic, ethnicity nonresponse) and race (American Indian or Alaska Native, Asian, Black, Native Hawaiian or other Pacific Islander, White, race non-response), allowing students to check as many boxes as they want (College Board, 2016). From these responses, a College Board "derived aggregate race/ethnicity" was re-created based on the U.S. Department of Education reporting guidelines that includes the following categories: no response; American Indian/Alaska Native; Asian; Black; Hispanic/Latino; Native Hawaiian or Other Pacific Islander; White; other; two or more races, non-Hispanic (College Board, 2016).[^5] Information about student socioeconomic status is limited. However, we create a measure of whether the prospect is first-generation college student based on their parents' level of education that was self-reported on the College Board questionnaire \[KS double check this\].

[^5]: Any student who selects a Hispanic ethnicity category is defined as Hispanic/Latino, regardless of the race categories they select. Additionally, non-Hispanic students who check “American Indian or Alaska Native” and another race group are defined as “two or more races, non-Hispanic.”

@@

**Research design**. We utilize a multiple, quantitative case study design in which metropolitan areas are cases. The research questions of this study aim to explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to generalize Geomarket variation nationally, but rather to explore how differences across Geomarkets landscapes shape prospect pools based on the geographical model of metropolitan areas and their demographic characteristics. Such aims align with the purpose of a multiple case study design to gain a deeper understanding of the phenomenon of interest by comparing and contrasting across different settings \[CITE YIN 2001\]. In doing so, multiple cases allow researchers to explore the "how" and "why" of the phenomenon in rich and nuanced contexts. We therefore focus on minimizing the number of metropolitan areas in order to provide sufficient analytic depth while also situating each case within its specific historic and contemporary geographical landscape \[TIE IN THEORETICAL FRAMEWORK HERE A BIT MORE\]. Online Appendix XXX presents results for a larger number of metropolitan areas.

Case selection was informed by based on purposeful sampling to identify “information-rich” cases (Patton, 2002; Yin, 2009) that can provide in-depth knowledge about the phenomenon of interest. We first considered metropolitan areas as information rich cases across two broad dimensions: 1) Geographical and urban model structures and 2) variation across racial/ethnic and socioeconomic characteristics. Based on our read of @RN4982, we are interested in "major metropolitan area\[s\] \ldots composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs" (p. 11-12). Building on our conceptual framework, we are interested in racial and socioeconomic inequality between geomarkets within a metropolitan area (RQ1) and how this inequality contributes to inequality in which prospective students are targeted by student list purchases (RQ2). \[DISCUSSION OF WHICH CASES TO SELECT. KARINA\]. Choice of metropolitan areas is informed by data availability. Although Census data are available for all metropolitan areas (RQ1) we do not have good candidate student list purchases for all metropolitan areas (RQ2).

Then, our purposeful sampling approach also took into consideration student list orders! \[SAY SOMETHING ABOUT WHAT IS A GOOD CANDIDATE STUDENT LIST PURCHASE -- PURCHASE WITH NOT TOO MANY CRITERIA THAT INCLUDE ALL GEOMARKETS IN A METRO\].
